---
title: "R for Data Science Chapter 3"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 3.2 First Steps 

Answer the question do cars with big engines use more fuel than cars with small
engines?

```{r}
mpg
```

Lets try plotting the cars efficiency (`hwy`) vs. it's car engine size in liters
(`displ`)

```{r hwy_vs_disp}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 3.2.4 Exercises

1. Run `ggplot(data = mpg)`. What do you see?

```{r, blankcanvas}
ggplot(data = mpg)
```

An empty canvas

2. How many rows are in `mpg`? How many columns?

```{r}
mpg
```

Looks like 234 rows and 11 columns

3. What does the drv variable describe? Read the help for ?mpg to find out

```{r}
?mpg
```

What type of drive train a car has.

4. Make a scatterplot of `hwy` vs. `cyl`

```{r hwy_vs_cyl}
ggplot(data = mpg, aes(y = hwy, x = cyl)) + geom_point()
```

5. What happens if you make a scatterplot of `class` vs. `drv`? Why is the plot
   not useful?

```{r class_vs_drv}
ggplot(data = mpg, aes(y = class, x = drv)) + geom_point()
```

It's not useful because you are plotting two different categorical variables
against each other. It might make more sense to represent the data as a count of
each combination of categories:

```{r classcount_by_drv}
ggplot(data = mpg, aes(x = class, fill = drv)) + geom_bar()
```


# 3.3 Aesthetic mappings

Can we explain the points that don't fit the general trend using another
categorical variable? Do we expect these values to be, say, hybrid cars?

Use aesthetics to change the color of the points to indicate the different
classes of cars.

```{r hwy_vs_displ_by_class}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

It looks like the outliers are all two-seaters. These are usually sports cars
which have big engines for small body sizes and therefore get slightly better
gas mileage.

We can also change the class to be marked by size. But this doesn't really make
sense does it? There is no value associated with being an SUV vs. a 2seater.
Thus, assigning a size doesn't really work and R warns us.
```{r hwy_vs_displ_by_class_size}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```


We can also change the transparency or shape of the points as well. However,
transparency falls under the same nonsense as size for this particular data.
```{r hwy_vs_displ_by_class_alpha}
# Left
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

```{r hwy_vs_displ_by_class_shape}
# Right
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

For the shapes R is warning us that more than 6 shapes is hard to discriminate
and therefore it drops the seventh category. It wants us to manually specify
shape instead. Which we haven't learned how to do yet.

We can also change the overall color of all the points by specifying it outside
the the aesthetic mapping
```{r hwy_vs_displ_colorblue}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

Other things you can change are
* Name of a color as a character string.
* Size of a point in mm
* Shape of a point as a number (R has 25 built in shapes)

## 3.3.1 Exercises

1. What's gone wrong with this code? Why are the points not blue?

```{r exercise_3-3-1_1}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

The color specification was put inside aesthetic instead of outside it, so the
function treated it as a categorical variable

2. Which variables in `mpg` are categorical? Which variables are continuous?
   (Hint: type `? mpg` to read the documentation for the data set). How can you
   see this information when you run `mpg`?

```{r}
mpg
```

Can see the datatype underneath the variable name. Chr are character values
which can be treated as categorical. Similarly int are integers which can be
treated either continuously or as categorical. dbl are float numbers which are
continuous.

3. Map a continuous variable to `color`, `size`, and `shape`. How do these
   aesthetics behave differently for categorial vs. continuous variables?

```{r exercise_3-3-1_3-1}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = hwy))
```

Color changes to a gradient instead of discrete colors

```{r, error=TRUE}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, shape = hwy))
```

Shape will not work

```{r exercise_3-3-1_3-2}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, size = hwy))
```

Size also gets mapped onto a continuous scale


4. What happens if you map the same variable to multiple aesthetics?

```{r exercise_3-3-1_4}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, size = hwy, color = hwy))
```

The value gets mapped to both aesthetics

5. What does the `stroke` aesthetic do? What shapes does it work with? (Hint:
   use `? geom_point`) The information about stroke can be found in
   `vignette("ggplot2-specs")` which was found after looking at the `?
   geom_point` documentation. IN there it says that stroke controls the outside
   of a shape only for shapes 21-24. Lets try setting the stroke.

```{r exercise_3-3-1_5}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy), shape = 21, stroke = 4)
```

6. What happens if you map an aesthetic to something other than a variable name,
   like `aes(colour = displ < 5)`? Note, you'll also need to specify x and y.

```{r exercise_3-3-1_6}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = displ < 5))
```

It takes the solution to displ < 5 as the categorical variable for the plotting

# 3.5 Facets

We could add additional variables with the aesthetics, but it might be better to
split the data into facets. Allowing us to plot the same thing over a series of
different categories.

Here is an example of plotting the `hwy` vs. `displ` for each class of vehicle

```{r hwy_vs_displ_by_class2}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

What if we want to facet on two different sets of variables?
Here we will split on the number of cylinders on the x and the type of
drive-train on the y.
```{r hwy_vs_displ_by_class_and_cyl}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ cyl)
```

## 3.5.1 Exercises

1. What happens if you facet on a continuous variable?

```{r}
mpg
```

```{r exercise_3-5-1_1}
ggplot(data = mpg, aes(x = cyl, y = hwy)) + geom_point() + facet_wrap(~ displ)
```

It still plots a plot for every single value.

2. What do the empty cells in a plot with `facet_grid(drv ~ cyl)` mean? 

```{r exercise_3-5-1_2-1}
ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point() + facet_grid(drv ~ cyl)
```

The empty plots mean you have no data for that location.

How do they relate to this plot?

```{r exercise_3-5-1_2-2}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = drv, y = cyl))
```

They are the locations you have no data.

3. What plots does the following code make? What does `.` do?

```{r exercise_3-5-1_3-1}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)
```

```{r exercise_3-5-1_3-2}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

The dot appears to say don't plot anything for that dimension.

4. Take the first faceted plot in this section:

```{r exercise_3-5-1_4}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

Advantages, allows you to seperate out each group and consider individually.
Disadvantages, can't see how well the groups fit in the data overall.
Balancing - why not plot both?

5. Read `?facet_wrap`. What does `nrow` do? What does `ncol` do?

`nrow` and `ncol` change the number of rows and columns respectively

What other options control the layout of the individual panels? 

`drop` allows you to drop categorical values that don't show up
`dir` allows you to switch between vertical and horizontal wrapping

Why doesn't `facet_grid()` have `nrow` and `ncol` arguments?

Because the rows and columns are set by your data.

6. When using `facet_grid()` you should usually put the variable with more
   unique levels in the columns. Why?

Screens are typically rectangular so you have more space wide then you do tall

# 3.6 Geometric objects

Can change the type of plot by changing the geom

```{r hwy_vs_displ_point}
#left
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy))
```

```{r hwy_vs_displ_smooth}
# right
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x= displ, y = hwy))
```

How can we change the aesthetics of the smooth fit?

```{r, hwy_vs_displ_by_drv_line}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

This is weird, what if we plotted this over top of the actual data that goes
with each drivetrain? How can we do that? We will find out soon.

First lets play with the grouping of data. In the plot above we changed the
linestyle for drivetrain giving us a distinguishing feature for each group. What
if we want all the groups to look the same?

```{r hwy_vs_displ}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

Now lets just group by drive train

```{r hwy_vs_displ_by_drv_grp}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

Lets instead change the color for each but don't put a legend

```{r hwy_vs_displ_by_drv_color}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```

Now lets plot the data with the overall fit:

```{r hwy_vs_displ_point_smooth}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

However, this introduces duplication in the code and could lead to mistakes when
one goes to change things. Lets instead put the mapping in the ggplot call

```{r hwy_vs_displ_point_smooth2}
ggplot(data = mpg, mapping = aes(x= displ, y = hwy)) +
  geom_point() +
  geom_smooth()
```

Now we can overwrite a local layer seperately from the overall layer

```{r hwy_vs_displ_by_class_color}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth()
```

Now what if we wanted to show the all the data but only the fit for the
subcompact class? We can feed only the subcompact data within geom_smooth but
keep everything else the same

```{r hwy_vs_displ_smooth_subcompact}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(color = class)) +
  geom_smooth(data = filter(mpg, class == "subcompact"), se = FALSE)
```

# 3.6.1 Exercises

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?

* `geom_line()`
* `geom_boxplot()`
* `geom_hist()`
* `geom_area()`

2. Run this code in your head and predict what the output will look like. Then,
   run the code in R and check your predictions.

```{r Exercise_3-6-1_2-1}
ggplot(mpg, aes(x = hwy, y = cty)) + geom_area()
```



```{r Exercise_3-6-1_2-2}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

Should draw loess lines for each individual class

3. What does `show.legend = FALSE` do?

doesn't show the legend for a layer

What happens if you remove it?

The legend shows up for that layer

Why do you think it was used earlier in the chapter?

To remove the legend.

4. What does the `se` argument to `geom_smooth()` do?

Removes the confidence interval estimate of the smooth line. Default of se is
a 0.95 confidence interval

5. Will these two graphs look different? Why/why not?

```{r Exercise_3-6-1_5}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

Yes but the second one is redundant.

6. Recreate the R code necessary to generate the following graphs.

```{r Exercise_3-6-1_6_1}
ggplot(data = mpg, aes(x = displ, y= hwy)) +
  geom_smooth(se = FALSE) + geom_point()
```

```{r Exercise_3-6-1_6_2}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point() +
  geom_smooth(aes(group = drv), color = "red", se = FALSE)
```

```{r Exercise_3-6-1_6_3}
ggplot(data = mpg, aes(x = displ, y = hwy, color = drv)) + geom_point() +
  geom_smooth(se = FALSE)
```

```{r Exercise_3-6-1_6_4}
ggplot(data = mpg, aes(x = displ, y = hwy,
              color = drv)) +
  geom_point() +
  geom_smooth(aes(color=NULL), se = FALSE)
```

```{r Exercise_3-6-1_6_5}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth(se = FALSE)
```


```{r Exercise_3-6-1_6_6}
ggplot(data = mpg, aes(x = displ, y = hwy,
                       color = drv)) +
  geom_point() +
  geom_smooth(aes(color = NULL, linetype = drv), se = FALSE)
```

```{r Exercise_3-6-1_6_7}
ggplot(data = mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(shape = 21, aes(fill = drv), color = "white", stroke =2, size =2)
```

# 3.7 Statistical transformations

Now lets look at bar charts as an intro to graphs that require manipulation of
the data

```{r diamonds_cut_bar}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut))
```

Where did count come from?
```{r}
diamonds
```

Its not in the diamonds dataset, its calculated from a stat. We can also use
`stat_count()` instead

```{r diamonds_cut_count}
ggplot(data = diamonds) +
  stat_count(mapping = aes(x = cut))
```

There are three reasons why you would want to use stat explicitly

1. Override the default stat

```{r diamonds_cut_freq}
demo <- tribble(
  ~cut, ~freq,
  "Fair", 1610,
  "Good", 4906,
  "Very Good", 12082,
  "Premium", 13791,
  "Ideal", 21551
)

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = "identity")
```

2. Might want to override the default mapping from transformed variables to
   aesthetics.

```{r diamonds_cut_stat_prop}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1))
```

3. Draw attention to the statistical transformation to summarize the y values
   for each x

```{r diamonds_depth_vs_cut}
ggplot(data = diamonds) +
  stat_summary(
    mapping = aes(x = cut, y = depth),
    fun.ymin = min,
    fun.ymax = max,
    fun.y = median
  )
```

## 3.7.1 Exercises

1. What is the default geom associated with `stat_summary()`? How could you
   rewrite the previous plot to use that geom function instead of the stat
   function?

```{r Exercise_3-7-1_1}
ggplot(data = diamonds, mapping = aes(x = cut, y = depth)) +
  geom_pointrange(stat = "summary", fun.y = median, fun.ymax = max, fun.ymin = min)
```

2. What does `geom_col()` do? How is it different than `geom_bar`

`geom_col()` is a shortcut for running `geom_bar` with the stat set to identity


3. Most geoms and stats come in pairs that are almost always used in concert.
   Read through the documentation and make a list of all the pairs. What do they
   have in common?

`stat_bin` with `geom_bar`
`stat_boxplot` with `geom_boxplot`
`stat_contour` with `geom_contour`

They are usually named the same and are used by each other.

4. What variable does `stat_smooth()` compute? What parameters control its
   behaviour?

It computes a predicted value for y, a ymin, ymax and a standard error. Several
parameters control its behavior

5. In our proportion bar chart, we need to set `group = 1`. Why? In other words
   what is the problem with these two graphs?

```{r Exercise_3-7-1_5-1}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop..))
```

Its calculating the proportion over all of the data

```{r Exercise_3-7-1_5-2}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = color, y = ..prop..))
```

```{r Exercise_3-7-1_5-3}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, y = ..prop.., fill = color, group = color))
```

Still calculating the proportion over all of the data.

# 3.8 Position adjustments

We can also add color to bar charts
```{r diamonds_cut_color_cut}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, color = cut))
```

```{r diamonds_cut_fill_cut}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = cut))
```

Now lets change the fill to some other variable
```{r diamonds_cut_fill_clarity}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

There are other positions we could specify other than the default `stack`.

```{r diamonds_cut_fill_clarity_posidentity}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) +
  geom_bar(alpha = 1/5, position = "identity")
```

```{r diamonds_cut_fill_clarity_posfill}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

```{r diamonds_cut_fill_clarity_posdodge}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

We can also adjust the position of other plots as well
```{r hwy_vs_displ_posjitter}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter")
```

## 3.8.1 Exercises
1. What is the problem with this plot? How could you improve it?

```{r Exercise_3-8-1_1-1}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_point()
```

```{r Exercise_3-8-1_1-2}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_jitter()
```

```{r Exercise_3-8-1_1-3}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_hex() + 
    scale_fill_gradient(trans = "log2")
```
2. What parameters to `geom_jitter()` control the amount of jittering?

width and heigth

3. Compare and contrast `geom_jitter()` with `geom_count()`

```{r Exercise_3-8-1_3}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) + geom_count()
```

Maps the size of the points to the number of observations

4. What's the default position adjustment for `geom_boxplot()`? Create
   a visualization of the `mpg` dataset that demonstrates it.

```{r Exercise_3-8-1_4-1}
ggplot(data = mpg, mapping = aes(x = drv, y= hwy, color = class)) + 
    geom_boxplot()
```

```{r Exercise_3-8-1_4-2}
ggplot(data = mpg, mapping = aes(x = as.factor(cty), y= hwy, color = class)) + 
    geom_violin()
```


# 3.9 Coordinate systems

Flipping the coordinate systems can be helpful.

```{r flip_coord}
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot()
ggplot(data = mpg, mapping = aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip()
```

Plotting maps has some special code for it
```{r map_nz1}
nz <- map_data("nz")

ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
```

```{r map_nz2}
ggplot(nz, aes(long, lat, group = group)) +
  geom_polygon(fill = "white", color = "black") +
  coord_quickmap()
```

Can switch over to polar coordinates
```{r diamonds_polar}
bar <- ggplot(data = diamonds) + 
  geom_bar(
    mapping = aes(x = cut, fill = cut), 
    show.legend = FALSE,
    width = 1
  ) + 
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)

bar + coord_flip()
bar + coord_polar()
```


## 3.9.1 Exercises
1. Turn a stacked bar chart into a pie chart using `coord_polar()`

```{r Exercise_3-9-1_1}
bar <- ggplot(data = diamonds) +
  geom_bar(
    # force stacked bar with x= factor(1)
    mapping = aes(x = factor(1), fill = cut),
    show.legend = FALSE,
    width =1) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)
bar
bar+coord_flip()
bar + coord_polar(theta = "y")
```

2. What does `labs()` do? Read the documentation.

Allows you to add labels to your plot. 

* title gives the title
* subtitle gives a title under the title
* caption is displayed at the bottom-right of the plot
* tag is displayed at the top left of the plot
* label allows for changing the axis labels

```{r Exercise_3-9-1_2}
bar <- ggplot(data = diamonds) +
  geom_bar(
    # force stacked bar with x= factor(1)
    mapping = aes(x = factor(1), fill = cut),
    show.legend = FALSE,
    width =1) +
  theme(aspect.ratio = 1) +
  labs(x = NULL, y = NULL)
bar + coord_polar(theta = "y") + 
  labs(title = "This is the title", subtitle = "Look! The subtitle",
       caption = "Data source can go here",
       tag = "A.")
```

3. What's the difference between `coord_quickmap()` and `coord_map()`?

`coord_map` projects the earth onto a flat 2d plane. `coord_quickmap` does a
quick approxmiation that preserves straight lines But works better for areas
near the equator

4. What does the plot below tell you about the relationship between city and
highway mpg? Why is `coord_fixed()` important? What does `geom_abline()` do?

```{r Exercise_3-9-1_4_1}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() +
  coord_fixed()
```

Lets try it without `coord_fixed()`
```{r Exercise_3-9-1_4_2}
ggplot(data = mpg, mapping = aes(x = cty, y = hwy)) +
  geom_point() +
  geom_abline() 
```

`coord_fixed` keeps the axes on a physical correspondence rather than scaling
them to be equal. `geom_abline` adds a line with an intercept and slope.

# 3.10 The layered grammar of graphics

Extendable language for any plot
